Coursera Reproducible Research Project 1
-------------------------------------------
**Introduction**  
The following is a brief exploratory data analysis of a dataset collected using acticity monitoring devices such as Fitbit, Jawbone etc. 

**Preparing the data**  
The first step is to download and unzip the data file and produce a data frame, as follows.
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
    destfile="activitydata.zip",method="curl")
unzip("activitydata.zip")
actdata<-read.csv("activity.csv")
```
**Question 1: Number of steps taken each day**  
To explore the average number of steps taken per day, a new vector of values is calculated using tapply.

```{r, echo=TRUE}
steps.day<-tapply(actdata$steps,actdata$date,sum,na.rm=TRUE)
summary(steps.day,digits=c(4,4,5,4,4,4))
```

The summary data show that the mean number of steps per day was `r round(mean (steps.day),0)` and the median number of steps was `r median(steps.day)`. This suggests that the distribution is negatively skewed, and this is  illustrated by the histogram below.

``` {r}
hist(steps.day,xlab="Number of steps per day",ylab="Freq.",
    main="Histogram: steps per day",breaks=10,col="lightyellow")
```

This shows a number of zero values, likely due to missing values in the data set (see later)

**Question 2: The daily routine**  
The time series plot below shows the daily routine in terms of the average number of steps per five minute interval (the x-axis shows the index number of the interval). 
``` {r} 
routine<-tapply(actdata$steps,actdata$interval,mean,na.rm=TRUE)
plot(routine,type="l",ylab="Average number of steps",
    xlab="5-minute interval (index)",main="Mean daily activity pattern",
    col="darkblue")
```

``` {r}
which(routine==max(routine))
max(routine)
```

The most active 5 minute interval of the day is index 104 (08:35 to 08:40), which averages 206 steps across the observed days.

**Question 3: Removing NA values**  
To explore whether missing values have a significant distorting effect on the data, the number of steps per day is recalculated, replacing all NA values in the data with the median value for the 5-minute interval to which it belongs.

``` {r}
index<-which(is.na(actdata$steps))
actdata.noNA<-actdata
for (i in index) {
    currentinterval<-actdata$interval[i]
    newvalue<-median(actdata$steps[actdata$interval==currentinterval],na.rm=TRUE)
    actdata.noNA$steps[i]<-newvalue
}
steps.day.2<-tapply(actdata.noNA$steps,actdata.noNA$date,sum)
```

The histogram of the corrected data is shown below.
``` {r}
hist(steps.day.2,xlab="Number of steps per day",ylab="Freq.",
    main="Histogram: steps per day, filled data",breaks=10,col="lightyellow")
```

Summary information for the data is as follows:

``` {r}
summary (steps.day.2,digits=c(4,4,5,4,4,4))
```

This shows that the summary values have not changed dramatically, though the zero values have disappeared and the mean has increased slightly.

**Question 4: Comparing weekday vs weekend**  
In order to compare daily weekday and weekend routines, the date field was recoded from a factor to a date format and the weekdays function used to assign a new variable showing whether the day falls on a weekday or a weekend.

``` {r}
actdata$date<-strptime(as.character(actdata$date),format="%F")
actdata$day<-weekdays(actdata$date)
actdata$weekday<-rep("weekday",nrow(actdata))
actdata$weekday[actdata$day=="Saturday"|actdata$day=="Sunday"]<-"weekend"
```

The mean activity per 5-minute interval was calculated again, first conditioning on whether the day was a weekday or weekend.

```{r}
routine.weekday<-with(subset(actdata,actdata$weekday=="weekday"),
    tapply(steps,interval,mean,na.rm=TRUE))
routine.weekend<-with(subset(actdata,actdata$weekday=="weekend"),
    tapply(steps,interval,mean,na.rm=TRUE))
```

Finally, plots were produced comparing the weekday and weekend routine. 

```{r fig.width=12}
# make the plots
par(mfrow=c(1,2))
plot(routine.weekday,type="l",col="darkblue",ylim=c(0,250),
    xlab="5-minute interval (index)",ylab="Average number of steps",
    main="Daily routine: weekday")
plot(routine.weekend,type="l",col="darkblue",ylim=c(0,250),
    xlab="5-minute interval (index)",ylab=NA,main="Daily routine: weekend")
```













